---
title: "Changing the axis positions"
author: "Claus O. Wilke"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 4
vignette: >
  %\VignetteIndexEntry{Changing the axis positions}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---
The **cowplot** package defines a function `switch_axis_position()` that allows us to take the x and/or y axis of a plot and copy or move it to the other side of the plot (y axis to right, x axis to top). Here are some examples of moving or copying various axes. As we can see, this works regardless of the chosen theme.

Importantly, the return object is a gtable, not a ggplot object, so we need to feed it into `ggdraw()` to display in the usual manner or to save with `ggsave()` or `save_plot()`.
```{r, message=FALSE, fig.width=4.25, fig.height=3.4}
require(cowplot)
require(grid) # for unit()
theme_set(theme_cowplot(font_size=12)) # reduce default font size
p1 <- ggplot(mtcars, aes(mpg, disp)) + geom_line(colour = "blue")
ggdraw(switch_axis_position(p1 + theme_gray(), axis = 'y'))
ggdraw(switch_axis_position(p1 + theme_bw(), axis = 'x', keep = 'x'))
ggdraw(switch_axis_position(p1, axis = 'xy', keep = 'xy'))
ggdraw(switch_axis_position(p1 + theme(axis.ticks.length=unit(0.3, "cm"),
        axis.ticks.margin=unit(0.2, "cm")), axis = 'xy'))
```

Moving the axes also works if there is a legend. What is not currently implemented, however, is moving axes on faceted plots.
```{r, message=FALSE, fig.width=4.25, fig.height=3.4}
p2 <- ggplot(diamonds, aes(clarity, fill = cut)) + geom_bar() +
  theme(axis.text.x = element_text(angle=70, vjust=0.5))
ggdraw(switch_axis_position(p2, axis = 'x'))
```

In combination with some clever gtable manipulations, we can also align multiple plots with axes that have been switched. Note that the function `gtable_squash_rows()` I'm using here is one of a few functions **cowplot** defines for manipulating gtables.
```{r, message=FALSE, fig.width=4.25, fig.height=5.1}
require(gtable)
# top plot
p1 <- ggplot(mtcars, aes(mpg, disp)) + geom_line(colour = 'blue') +
  background_grid(minor='none')
g1 <- switch_axis_position(p1, 'xy') # switch both axes
g1 <- gtable_squash_rows(g1, length(g1$height)) # set bottom row to 0 height

p2 <- ggplot(mtcars, aes(mpg, qsec)) + geom_line(colour = 'green') + ylim(14, 25) +
  background_grid(minor='none')
g2 <- ggplotGrob(p2)
g2 <- gtable_add_cols(g2, g1$widths[5:6], 4) # add the two additional columns that g1 has
g2 <- gtable_squash_rows(g2, 1:2) # set top two rows to 0 height
plot_grid(g1, g2, ncol=1, align='v')
```
