---
title: "Mixing different plotting frameworks"
author: "Claus O. Wilke"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Mixing different plotting frameworks}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---
```{r message = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  fig.width = 4,
  fig.height = 3
)
```

All cowplot functions that take plot objects as input (`ggdraw()`, `draw_plot()`, `plot_grid()`) can handle several different types of objects in addition to ggplot2 objects. Most importantly, they can handle plots produced with base R graphics. However, this functionality is only available if you have the package [gridGraphics](https://CRAN.R-project.org/package=gridGraphics) installed.

As the first example, we draw a base graphics plot with `ggdraw()` and style the background with the ggplot2 themeing mechanism.

```{r message = FALSE}
library(ggplot2)
library(cowplot)

# define a function that emits the desired plot
p1 <- function() {
  par(
    mar = c(3, 3, 1, 1),
    mgp = c(2, 1, 0)
  )
  boxplot(mpg ~ cyl, xlab = "cyl", ylab = "mpg", data = mtcars)
}

ggdraw(p1) +
  theme(plot.background = element_rect(fill = "cornsilk"))
```

We can also add a logo to the plot.
```{r}
logo_file <- system.file("extdata", "logo.png", package = "cowplot")

ggdraw() + 
  draw_image(
    logo_file,
    x = 1,
    width = 0.13, height = 0.2,
    hjust = 1
  ) +
  draw_plot(p1)
```

And we can draw base graphics and ggplot2 graphics side-by-side in a plot grid.

```{r fig.width = 6}
p2 <- ggplot(data = mtcars, aes(factor(cyl), mpg)) + geom_boxplot()
plot_grid(p1, p2)
```

Base R plots can be stored in the form of functions that emit the desired plots (demonstrated above), as recorded plots, or using a convenient formula interface.

To create a recorded plot, we first draw the base plot, then we record it with `recordPlot()`, and then we can draw it with `ggdraw()`.

```{r}
# create base R plot
par(mar = c(3, 3, 1, 1), mgp = c(2, 1, 0))
boxplot(mpg ~ cyl, xlab = "cyl", ylab = "mpg", data = mtcars)
# record previous base R plot and then draw through ggdraw()
p1_recorded <- recordPlot()
ggdraw(p1_recorded)
```

We can store arbitrarily complex plotting code in formulas by enclosing it into curly braces.

```{r}
# store base R plot as formula
p1_formula <- ~{
  par(
    mar = c(3, 3, 1, 1), 
    mgp = c(2, 1, 0)
  )
  boxplot(mpg ~ cyl, xlab = "cyl", ylab = "mpg", data = mtcars)
}
ggdraw(p1_formula)
```

There is also support for lattice graphics and grid grobs.

```{r message = FALSE, fig.width = 8, fig.height = 6}
# base R
p1 <- ~{
  par(
    mar = c(3, 3, 1, 1), 
    mgp = c(2, 1, 0)
  )
  boxplot(mpg ~ cyl, xlab = "cyl", ylab = "mpg", data = mtcars)
}

# ggplot2
p2 <- ggplot(data = mtcars, aes(factor(cyl), mpg)) + geom_boxplot()

# lattice
library(lattice)
p3 <- bwplot(~mpg | cyl, data = mtcars)

# elementary grid graphics objects
library(grid)
p4 <- circleGrob(r = 0.3, gp = gpar(fill = "skyblue"))

# combine all into one plot
plot_grid(p1, p2, p3, p4, rel_heights = c(.6, 1), labels = "auto")
```

Other packages are supported as long as they return grid grobs.

```{r message = FALSE, fig.width = 4.5, fig.height = 4.5}
library(VennDiagram)

p_venn <- draw.pairwise.venn(
  100, 70, 30,
  c("First", "Second"),
  fill = c("light blue", "pink"),
  alpha = c(0.7, 0.7),
  ind = FALSE
)

# plot venn diagram and add some margin and enclosing box
ggdraw(p_venn) +
  theme(
    plot.background = element_rect(fill = NA),
    plot.margin = margin(12, 12, 12, 12)
  )
```
