---
title: "Mixing different plotting frameworks"
author: "Claus O. Wilke"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 4
vignette: >
  %\VignetteIndexEntry{ixing different plotting frameworks}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

Any cowplot functions that take plot objects as input (`ggdraw()`, `draw_plot()`, `plot_grid()`) can handle several different types of objects in addition to ggplot2 objects. Most importantly, they can handle plots produced with base R graphics.

```{r message = FALSE, warning = FALSE}
library(ggplot2)
library(cowplot)

# define function that emits desired plot
p1 <- function() {boxplot(mpg ~ cyl, xlab = "cyl", ylab = "mpg", data = mtcars)}

ggdraw(p1) +
  theme(plot.background = element_rect(fill = "cornsilk"))

ggdraw() +
  draw_image("http://jeroen.github.io/images/tiger.svg", scale = 0.9) +
  draw_plot(p1)

p2 <- ggplot(data = mtcars, aes(factor(cyl), mpg)) + geom_boxplot()
plot_grid(p1, p2)
```

Base R plots can be stored in the form of functions that emit the desired plots, as recorded plots, or using a convenient formula interface.

```{r}
# create base R plot
boxplot(mpg ~ cyl, xlab = "cyl", ylab = "mpg", data = mtcars)
# record previous base R plot and then draw through ggdraw()
p1_recorded <- recordPlot()
ggdraw(p1_recorded)
```

```{r}
# store base R plot as formula
p1_formula <- ~boxplot(mpg ~ cyl, xlab = "cyl", ylab = "mpg", data = mtcars)
ggdraw(p1_formula)
```


```{r message = FALSE, fig.width = 8, fig.height = 6}
library(ggplot2)
library(lattice)
library(cowplot)

# base R (note formula format for base graphics)
# package "gridGraphics" needs to be installed for this to work
p1 <- ~{
  par(cex = .8)
  boxplot(mpg ~ cyl, xlab = "cyl", ylab = "mpg", data = mtcars)
}

# lattice
p2 <- bwplot(~mpg | cyl, data = mtcars)

# ggplot2
p3 <- ggplot(data = mtcars, aes(factor(cyl), mpg)) + geom_boxplot()

# combine all into one plot
plot_grid(p1, p2, p3, rel_heights = c(1, .6), labels = "auto")
```

```{r}
ggdraw(p1)
```


```{r fig.with = 8}
p_sqrt <- ~plot(sqrt)
p_volcano <- ~image(volcano)

plot_grid(p_sqrt, p_volcano, labels = "auto")
```


```{r}
library(VennDiagram)

p_venn <- draw.pairwise.venn(100, 70, 30, c("First", "Second"), fill = c("light blue", "pink"), alpha = c(0.7, 0.7), ind = FALSE)

ggdraw(p_venn)
```
