---
title: "Themes"
author: "Claus O. Wilke"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 4
vignette: >
  %\VignetteIndexEntry{Themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

The cowplot package provides a consistent set of plot themes that cover a wide range of use cases. To see these themes in action, check out the book [*Fundamentals of Data Visualization,*](https://serialmentor.com/dataviz) which uses them throughout. However, note that the figures in the book use the font Myriad Pro. Since this font is not freely available, the cowplot themes use the default R font, which for most graphic devices is Helvetica. All themes have a `font_family` argument that you can use to customize the font for your own use.

## Basic use of plot themes

The theme `theme_half_open()` (or equivalently, `theme_cowplot()`) provides a classical plot appearance with two axis lines and no background grid. This theme works for most types of graphs, but it is most appropriate for scatter plots and line graphs.

```{r message = FALSE}
library(ggplot2)
library(cowplot)

p <- ggplot(mtcars, aes(disp, mpg)) + geom_point()

p + theme_half_open() # identical to theme_cowplot()
```

If you like a background grid, you can add it via the standard ggplot2 theme options. Alternatively, you can use the cowplot function `background_grid()`. This function needs to be placed after the theme call because the theme call overrides all previous theme settings.

```{r}
p +
  theme_half_open() +
  background_grid() # always place this after the theme
```

If you prefer a minimal appearance with grid and without axis lines, use `theme_minimal_grid()`.

```{r}
p + theme_minimal_grid()
```

All themes can be further customized. For example, the default font size is 14 point, which works well for figures that are 5 to 8 inches wide. For smaller figures, you may want to use 12 point. Also, we can modify the color of the grid lines for `theme_minimal_grid()`.

```{r}
p +
  theme_minimal_grid(
    font_size = 12,
    color = "grey70"
  )
```

## Choosing themes in the context of the plot

When choosing a plot theme, you should always pay attention to how it works for your specific plot. For example, if you are making a bar plot, `theme_half_open()` generates awkward floating bars because of ggplot2's automatic axis expansion (i.e., the x axis line does not lie at *y* = 0).

```{r}
p <- ggplot(mtcars, aes(factor(cyl))) +
  geom_bar(fill = "#56B4E9", alpha = 0.8) +
  theme_half_open()

stamp_ugly(p)
```

We can fix this problem by suppressing the axis expansion at the lower end.

```{r}
p <- ggplot(mtcars, aes(factor(cyl))) +
  geom_bar(fill = "#56B4E9", alpha = 0.8) +
  scale_y_continuous(
    # don't expand y scale at the lower end
    expand = expand_scale(mult = c(0, 0.05))
  )

p + theme_half_open()
```

Bar plots also tend to look awkward on background grids, and I would recommend against this use.

```{r}
stamp_ugly(
  p + theme_minimal_grid()
)
```

Instead, you can use `theme_minimal_hgrid()`, which only draws horizontal grid lines.

```{r}
p + theme_minimal_hgrid()
```

If you make a flipped bar plot with horizontal bars, you may instead want vertical grid lines, which you obtain with `theme_minimal_vgrid()`.

```{r}
p +
  coord_flip() +
  theme_minimal_vgrid()
```

## Faceting

You need to pay special attention when faceting. A theme may look just fine for a single plot but work poorly in a faceted plot. As an example, consider what happens when we facet with `theme_half_open()`.

```{r fig.width = 6.5}
p <- ggplot(mtcars, aes(disp, mpg)) +
  facet_wrap(~factor(cyl)) +
  geom_point() +
  theme_half_open(12)

stamp_ugly(p)
```

The three facets don't visually separate, because ggplot2 does not repeat the y axis line for each facet. We can improve the situation by drawing a border around each panel, either via standard ggplot2 theme calls or using the convenience function `panel_border()`. As with `background_grid()`, this function always has to be placed after the theme.

```{r fig.width = 6.5}
p + panel_border()  # always place this after the theme
```

In this particular case, maybe using both a panel border and a background grid is the best option.

```{r fig.width = 6.5}
p + 
  panel_border() +
  background_grid()
```

## Empty themes

The two themes `theme_map()` and `theme_nothing()` provide stripped-down themes without axes. `theme_map()` is similar to `theme_void()` from ggplot2 in that it retains the plot title, subtitle, caption, and legends, and simply removes axis ticks, lines, labels, and gridlines. All settings are matched to the other cowplot themes, so that you can mix plots using `theme_map()` and the other cowplot themes and they will look consistent.

To demonstrate how these themes work and how they interact with the other themes provided, let's first make a standard scatter plot with title, subtitle, caption, and legend.

```{r}
p <- ggplot(mtcars, aes(disp, mpg, color = wt)) +
  geom_point() +
  labs(
    title = "Fuel efficiency in cars",
    subtitle = "Bigger and more powerful cars are less efficient",
    caption = "Motor Trend, 1974"
  )

p + theme_cowplot(12)
```

The same plot looks as follows with `theme_map()`.
```{r}
p + theme_map(12)
```

Notice how all elements that have been retained look the same between this plot and the previous one.

If we draw the same plot with `theme_nothing()`, then everything except the plot panel is removed, including titles and legends.
```{r}
p + theme_nothing()
```

This theme is used by `ggdraw()` and `plot_grid()` for the enclosing drawing canvas.
