---
title: "Introduction to COWplot"
author: "Claus O. Wilke"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 4
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---
The **cowplot** package is a simple add-on to **ggplot2**. It is meant to provide a publication-ready theme for **ggplot2**, one that requires a minimum amount of fiddling with sizes of axis labels, plot backgrounds, etc. Its primary purpose is to give my students and postdocs an easy way to make figures that I will approve of. Thus, this package meets my personal needs and tastes. Yours may be different.

In addition to providing a modified plot theme, this package also offers functionality for custom annotations to **ggplot2** plots. It turns out that the easiest way to offer this functionality was to implement a general-purpose drawing canvas on top of **ggplot2**. As a result, you can achieve quite unusual effects with this package (see more below).  

The **cowplot** source code is available on github: [https://github.com/wilkelab/cowplot](https://github.com/wilkelab/cowplot)

# Plot design

The default **ggplot2** design is not particularly elegant, in my opinion. In particular, the gray background grid is a prime example of [chartjunk,](http://en.wikipedia.org/wiki/Chartjunk) which stands for graphical elements that distract the viewer without providing any useful information. For example, see this **ggplot2** visualization of the `mpg` data set:
```{r echo=FALSE, message=FALSE}
require(ggplot2)
```
```{r message=FALSE}
ggplot(mpg, aes(x = cty, y = hwy, colour = factor(cyl))) + 
   geom_point(size=2.5) + theme_gray()
```

The default design of **cowplot**, by contrast, is clean and simple. It looks similar to **ggplot2**'s `theme_classic()`, but there are important but subtle difference we'll discuss later.
```{r echo=FALSE, message=FALSE}
require(cowplot)
theme_set(theme_cowplot(font_size=11)) # default fontsize doesn't work well for online viewing
ggplot(mpg, aes(x = cty, y = hwy, colour = factor(cyl))) +
  geom_point(size=2.5)
```

In particular, the **cowplot** default theme works nicely in conjunction with the `save_plot()` function the package provides, such that the output pdfs are nicely formatted and scaled and don't require any additional parameter fiddling.
```{r eval=FALSE}
require(cowplot)
plot.mpg <- ggplot(mpg, aes(x = cty, y = hwy, colour = factor(cyl))) + 
  geom_point(size=2.5)
save_plot("mpg.pdf", plot.mpg) # use this instead of ggsave() when using cowplot
```

# Plot annotations

The other headache with **ggplot2** is adding labels and other annotations to a plot. **ggplot2** strictly separates the *plot panel* (the part inside the axes) from the rest of the plot, and it's generally easy to modify one or the other but not both.

To solve this issue in a generic way, **cowplot** implements a generic drawing layer on top of **ggplot2**. In this drawing layer, you can add arbitrary graphical elements on top of a graph. For example, let's take the mpg image from the previous section, label it with an A in the top-left corner, and mark it as a draft:
```{r message=FALSE}
plot.mpg <- ggplot(mpg, aes(x = cty, y = hwy, colour = factor(cyl))) + 
  geom_point(size=2.5)
ggdraw(plot.mpg) + 
  draw_plot_label("A", size = 13) + 
  draw_text("DRAFT!", angle = 45, size = 80, alpha = .2)
```

The function `ggdraw` sets up the drawing layer, and functions that are meant to operate on this drawing layer all start with `draw_`. The resulting object is again a standard **ggplot2** object, and you can do with it whatever you might do with a regular **ggplot2** plot, such as save it with `ggsave()`. [However, as mentioned before, I recommend using `save_plot()` instead. It's a thin wrapper around `ggsave()` that has some useful features.]

In fact, because `ggdraw()` produces a standard **ggplot2** object, we can draw on it with standard geoms if we want to. For example:
```{r message=FALSE}
t <- (0:1000)/1000
spiral <- data.frame(x = .45+.55*t*cos(t*15), y = .55-.55*t*sin(t*15), t)
ggdraw(plot.mpg) + 
  geom_path(data = spiral, aes(x = x, y = y, colour = t), size = 6, alpha = .4)
```

I don't know if this is useful in any way, but it shows the power of the approach.

Importantly, though, in all cases discussed so far, the main plot was below all other elements. Sometimes, you might want the plot on top. In this case, you can initialize an empty drawing canvas by calling `ggdraw()` without any parameters. You then place the plot by calling `draw_plot()`. Notice the difference in the two plots produced by the following code:
```{r message=FALSE, fig.show="hold"}
boxes <- data.frame(
  x = sample((0:36)/40, 40, replace = TRUE),
  y = sample((0:32)/40, 40, replace = TRUE)
)
# plot on top of annotations
ggdraw() + 
  geom_rect(data = boxes, aes(xmin = x, xmax = x + .15, ymin = y, ymax = y + .15),
            colour = "gray60", fill = "gray80") +
  draw_plot(plot.mpg) +
  draw_text("Plot is on top of the grey boxes", x = 1, y = 1,
            vjust = 1, hjust = 1, size = 10, fontface = 'bold')
# plot below annotations
ggdraw(plot.mpg) + 
  geom_rect(data = boxes, aes(xmin = x, xmax = x + .15, ymin = y, ymax = y + .15),
            colour = "gray60", fill = "gray80") + 
  draw_text("Plot is underneath the grey boxes", x = 1, y = 1,
            vjust = 1, hjust = 1, size = 10, fontface = 'bold')
```

Note that placing a plot on top of annotations only makes sense if the plot background is transparent. This is one of the main differences between `theme_cowplot()` and `theme_classic()`. If you tried the same example with `theme_classic()`, the gray boxes underneath the plot would not show.

Finally, the `draw_plot()` function also allows us to place graphs at arbitrary locations and at arbitrary sizes onto the canvas. This is useful for combining subplots into a layout that is not a simple grid, e.g. with one sub-plot spanning the entire width of the figure and two other figures using up half of the figure width:
```{r message=FALSE, fig.width=7, fig.height=5}
plot.iris <- ggplot(iris, aes(Sepal.Length, Sepal.Width)) + 
  geom_point() + facet_grid(. ~ Species) + stat_smooth(method = "lm") +
  background_grid(major = 'y', minor = "none") + # add thin horizontal lines 
  panel_border() # and a border around each panel
plot.diamonds <- ggplot(diamonds, aes(clarity, fill = cut)) + geom_bar() +
  theme(axis.text.x = element_text(angle=70, vjust=0.5))
ggdraw() +
  draw_plot(plot.iris, 0, .5, 1, .5) +
  draw_plot(plot.mpg, 0, 0, .5, .5) +
  draw_plot(plot.diamonds, .5, 0, .5, .5) +
  draw_plot_label(c("A", "B", "C"), c(0, 0, 0.5), c(1, 0.5, 0.5), size = 15)
```
The functions `background_grid()` and `panel_border()` are convenience functions defined by **cowplot** to save some typing when manipulating the background grid and panel border. 


Of course, we can also go crazy:
```{r message=FALSE, fig.width=7, fig.height=5}
ggdraw() +
  #geom_rect(data = boxes, aes(xmin = x, xmax = x + .15, ymin = y, ymax = y + .15),
  #          colour = "gray60", fill = "red", alpha=.03) +
  geom_path(data = spiral, aes(x = x, y = y, colour = t), size = 6, alpha = .4) +
  draw_plot(plot.diamonds, -.05, -.1, .55, .55) +
  draw_plot(plot.diamonds, .65, .4, .5, .5) +
  draw_plot(plot.mpg, .3, .3, .4, .4) +
  draw_plot(plot.iris, 0, .7, .7, .35 ) +
  draw_plot(plot.iris, .45, .0, .6, .3 )
```
